Learning Discrete-valued Bayesian Networks from Mixed Data

### 2015-11-13 BN_discretizer_paper.tex
misc updates

\section{Introduction}
\label{intro}
Bayesian networks \citep{Pearl_1988, PGM_2009} are an increasingly popular method for modeling uncertainty and causality in science and engineering. They provide an efficient factorization of the joint probability distribution over a set of random variables. Bayesian networks first emerged from artificial intelligence research and have been applied to a wide variety of problems, ranging from decision-making systems \citep{DMU_2015} to medical diagnoses~\todo{[cite]}. In most cases, we assume that all random variables in Bayesian networks are discrete, since many Bayesian network learning algorithms are unable to deal with continuous variables efficiently. However, many applications
require the use of continuous variables, such as position and velocity in dynamic systems.

There are two common approaches to extending Bayesian networks to continuous variables. The first is to model the conditional probability density of each continuous variable using specific families of parametric distributions, and then to redesign Bayesian network learning algorithms based on their parameters. One successful example is belief propagation in Gaussian graphical models \citep{Weiss_2011}. For other particle shapes \citep{Ihler_2009} and non-linear functions \tim{cite if you can}, these redesigned algorithms are computationally expensive and do not perform well.
\tim{Not sure what `particle shapes' are. Do you mean other parametric distributions?}

The second approach is discretization. Automated discretization methods have been studied in machine learning and statistics for many years \citep{Dougherty_1995, Kerber_1992, Holte_1993, Fayyad_1993}. Most discretization methods were designed for classification problems. They search for the best discretization policy\tim{maybe `scheme' is a better word than `policy', which in machine learning means something else.} of a continuous attribute by considering its interation with the target class variable. These discretization methods do not apply to continuous variables in Bayesian networks, where interations and dependencies between variables are more complicated. Discretization of a continuous variable in a Bayesian network requires considering the impact on the entire network. There is prior work on discretizing continuous variables in naive Bayesian networks and tree-augmented networks \citep{Fried_naive}, but only a few discretization methods for general Bayesian networks have been proposed \citep{Friedman_1996, Kozlov_1997, Monti_1998, Steck_2007}.

The discretization technique proposed by \citet{Friedman_1996} is the most well-known of these methods. It is based on the minimum description length principle (MDL): the optimal discretization policy minimizes the description length of the Bayesian network and discretization scheme. The MDL algorithm has a $\bigo\paren{N^3 + \paren{n_c {v_\text{max}}^{(n_c^p)_\text{max}} + {v_\text{max}}^{n_p}} \cdot N^2}$ runtime on Bayesian networks with a single continuous variable and multiple discrete variables, where $N$ is the number of data samples from the joint distribution, $n_p$ and $n_c$ are the numbers of parent and children variables for the continuous variable, $v_\text{max}$ is the largest cardinality over all variables in the Markov blanket, and  ${(n_c^p)_\text{max}}$ is the largest number of parent variables of the continuous variable's children. The MDL method iterates over all continuous variables until convergence in the case of multiple continuous variables, converging in a few cycles according to our tests of real-world data. Unfortunately, MDL tends to produce discrete variables with low cardinality.
\tim{Note: use $\text{thing}$ for subscripts such as $n_\text{max}$.}

\tim{Motivate / introduce Bayesian discretization methods perhaps? This would be a good place to cite Boulle and Lustgarten}

\tim{
The commented material below is good, but feels like too much detail for the intro. Is it possible to move it to later sections?
I recommend having one paragraph that starts with `In this paper we propose \ldots' and then lists the specific contributions. (Bayesian discretization for arbitrary networks, a faster algorithm, demonstrate that performance is preserved, etc.). Then we should have the final paragraph below that discusses organization.}

% In this paper, we propose a new Bayesian discretization method for continuous variables in Bayesian networks. This method extends the Bayesian single-variable discretization methods from \citet{Boulle_2006} and \citet{Lustgarten_2011}. We begin our method with the assumption that only one variable in a given Bayesian network is continuous and other variables are discrete. Furthermore we assume that the network structure is known in advance. Under this situation, we look for the most probable discretization policy $M$ given the data $D$ on other discrete variables. That is to say, the desired policy is $\textrm{arg}_M P(M \mid D)$. With Bayes' rule, $P(M \mid D) $ can be rewritten as $P(M) \cdot P(D \mid M) / P(D)$ , which is proportional to $P(M)\cdot P(D \mid M)$. Usually, $P(D \mid M)$ increases as the number of discretized intervals increases, since more intervals can provide more accuracy. $P(M)$, on the other hand, is designed to decrease as the number of intervals rises. As a result, we can automatically determine the number of intervals after discretization by maximizing  $P(M)\cdot P(D \mid M)$. In addition, the $P(D \mid M)$ can be factorized according to Bayesian network structure. With the proposed priors, we are able to restrict the number of discretized intervals so that it will not exceed the largest cardinality of variables in the Markov blanket by too much. This is important, since for most algorithms on Bayesian networks, their running times exponentially depend on the cardinality of variables. Another advantage of our method is the running time for learning. If there is only one continuous variable, the running time is $O(n_c  {v_\text{max}} \cdot N^2 + {v_\text{max}}^{n_p} \cdot N^2)$, which is significantly shorter than the running time of MDL principle discretization \citep{Friedman_1996}. With the approximation proposed in this paper, we can further reduce running time to $O({v_\text{max}}(n_p+n_c)N^2)$.

% For a Bayesian network with multiple continuous variables, we apply the discretization method discussed above iteratively. In the beginning, we prediscretize each continuous variable by equal-width method. The number of intervals for equal-width discretization is equal to largest cardinality of discrete variables. Then we iterate the one-variable discretization on each continuous variable in the following order: from the variable with highest topological order (leaves) to the variable with lowest topological order (root). Each time we finish a discretization procedure on one variable, we store the discretization result for later iterations. Experiments on real-world data show that with the same iteration order, our discretization method provides better discretization results than MDL principle method in terms of likelihood. The MDL principle method is easily stuck at local minima and also discretizes variables into too few intervals.

% Finally, we can combine our new discretization method with the K2 structure learning alorithm \citep{K2}. We first prediscretize all continuous variables before the K2 learning, since the K2 algorithm requires all variables to be discrete. Each time an edge is added in the K2 algorithm procedure, we rediscretize the continuous variables based on the updated Bayesian network and store the result for the next K2 algorithm iteration. By this principle, we are able to learn a discrete-valued Bayesian network from mixed data.

The rest of the paper is organized as follows: related work, including MDL principle discretization \citep{Friedman_1996} and MODL discretization \citet{Boulle_2006} are summarized in Section \ref{sec:related_work}. Preliminaries and notation are covered in Section \ref{sec:preliminaries}. The proposed single-variable discretization method is derived in Section \ref{sec:single_var} and is extended to Bayesian networks with multiple continuous variables and combined with Bayesian structure learning in Section \ref{sec:multi_var}. The paper concludes with the comparison of the proposed methods against prior work on real-world data in Section \ref{sec:experiments}.

@@

\section{Related Work}
\label{sec:related_work}
This section reviews MDL principle discretization \citep{Friedman_1996} and MODL discretization \citep{Boulle_2006}. The former is the predominant discretization method for continuous variables in Bayesian networks, and is compared to our proposed method in Section \ref{sec:experiments}. The latter is a discretization method for one continuous variable against a discrete target class. Our proposed method generalizes the MODL Bayesian discretization approach to Bayesian networks. The asymptotical equivalence between the MDL and MODL methods were examined in \citep{VL_2000}.

\subsection{MDL Principle Discretization}
The MDL principle was first proposed by \cite{MDL_1978}, and states that the best model for a dataset is one which minimizes the amount of information needed to describe it \citep{Grunwald_2009}. MDL methods trade off goodness-of-fit against model complexity to avoid overfitting. In the context of Bayesian networks, \cite{Friedman_1996} applied the MDL principle to determine the optimal number of discretization intervals for continuous variables and optimal the positions of their discretization boundaries. Their approach selects a discretization policy that minimizes the sum of the description length of discretized Bayesian network and the description length of the information necessary for recovering the original continuous values from the discretized data. In the single continuous variable case, the objective function is
\tim{Do you need double vertical bars? If we use single bars it might fit without having to use the small font size.}

@@

\noindent
where $I(\boldsymbol{A},\boldsymbol{B}) = \sum_{\boldsymbol{a},\boldsymbol{b}} \hat{P}_D (\boldsymbol{a},\boldsymbol{b}) \log {{\hat{P}_D (\boldsymbol{a},\boldsymbol{b})}\over{\hat{P}_D (\boldsymbol{a}),\hat{P}_D (\boldsymbol{b})}}$ \tim{consider deleting the right half of equation} is the mutual information between two discrete variables $\boldsymbol{A}$ and $\boldsymbol{B}$, $\|\boldsymbol{X}\|$ is the cardinality of discrete variable $\boldsymbol{X}$, $X^*_i$ is the discretized version of continuous variable $X_i$ and $H(p) = -p \log(p) - (1-p) \log(1-p)$. The optimal discretization policy is found with dynamical programming. If there are multiple continuous variables in a Bayesian network, we apply the algorithm to one variable at a time and iterate over all continuous variables. While iterating, only one variable is treated as continuous and other continuous variables are discretized based on an initial discretization scheme or the discretization result from a previous iteration.

The network structure is not always known in advance. Bayesian network structure learning can be extended to distributions including continuous variables by alternating between traditional discrete structure learning and optimal discretization. Starting with some preliminary discretization scheme, one applys a structure learning algorithm to identify the locally optimal graph structure. The discretization scheme is refined based on the learned network. The cycle is continued until convergence.

@@

The MODL method uses dynamic programming to find the optimal discretization scheme, and has a $\bigo\paren{N^3 + \|\text{Class}\| \cdot N^2}$ runtime, where $N$ is the of number of data instances. \cite{Lustgarten_2011} suggest several formulations for the prior term $P(\text{Model})$\tim{Did I state this correctly?}. If the prior is chosen to reflect uniform prior probability over discretizations the runtime can be reduced to $\bigo\paren{\|\text{Class}\| \cdot N^2}$.\tim{Is a sacrifice made? It is sub-optimal, right?}

@@

Samples can be drawn from the joint distribution represented by a Bayesian network by forward sampling.
Variables are sampled in topological order such that parental instantiations are always known~\citep[see][chap.~22]{algo_2009}.
Furthermore, the parent\rightarrow child sampling order can be reversed if the marginal probability of child variable $P(X_i)$ is known.
By Bayes' rule,

@@

Four prior principles adapted from the MODL approach define $P(M)$ and $P(D \mid M)$:
\begin{enumerate}
\item A discretization edge at $(x_i + x_{i+1})/2$ has probability
\begin{equation}
1 - {\exp(- L \cdot {{x_{i+1} - x_i}\over{x_N - x_1}})},
\end{equation}
where $L$ is the largest cardinality of discrete variables in $X$'s Markov blanket.
\item For a given interval of $M$, every distribution of $\Pi_X$ is equiprobable.
\item For each pair $(C_i,\boldsymbol{S_i})$ and a given interval of $M$, every distribution over $C_i$ given $\boldsymbol{S_i}$ is equiprobable.
\item The distributions over each $C_i$ and $\Pi_X$ in each interval of $M$ given each instantiation of $\boldsymbol{S_i}$ are independent.
\end{enumerate}

The prior term $P(M)$ has the form:

\begin{equation}
\label{eq:p_M}
P(M) = \prod_{i=1}^k \brock{ [1 - {\exp(- L \cdot {{x_{s_i+1} - x_{s_i}}\over{x_N - x_1}})}] \cdot \exp(L \cdot {{x_{s_{i}} - x_{s_{i-1} + 1}}\over{x_N - x_1}}) }\text{,}
\end{equation}

\noindent
where $s_i = \sum_{j=1}^k n_j$, $s_k = N$, and $s_0 = 0$. Before evaluating $P(D \mid M)$, note that

\begin{equation}
\label{eq:p_D_given_M}
P(D \mid M) \propto P(D_{\Pi_x} \mid M) \cdot \prod_{i = 1}^{n_c} P(D_{C_i} \mid M, D_{\boldsymbol{S_i}})\text{.}
\end{equation}

The graph structure allows us to only consider the interations between $X$ and the other variables in its Markov blanket, thus allowing for convenient factorization. For example, $P(D \mid M)$ for Figure \ref{fig:one} factors according to
\begin{equation}
P(D_{ P_1,P_2,P_3 } \mid M) \cdot P( D_{ C_1 } \mid M,D_{\boldsymbol{S_1}}) \cdot P(D_{C_2} \mid M,D_{ S_2  })\text{.}
\end{equation}

\begin{figure}[ht]
    \begin{tabular}{cc}
      \input{graph1}
    \end{tabular}
  \caption{Factorization of $P(D \mid M)$}
  \label{fig:one}
\end{figure}

The factorization arises from the conditional independencies in the Bayesian network.
Given a distribution over $X$ and a specific instance $x$, one can directly determine the conditional distribution over the parents $\Pi_x$.
The parents given $x$ are dependent, in general one cannot further factorize $P(D_{\Pi_x} \mid M)$.
Similarly, a distribution over $C_i$ is fully defined given $x$ and $\boldsymbol{S_i}$.
The children are independent given $X$, and thus factorize according to $\prod_{i = 1}^{n_c} P(D_{C_i} \mid M, D_{\boldsymbol{S_i}})$.

It is thus necessary to formulate $P(D_{\Pi_x} \mid M)$  and $P(D_{C_i} \mid M, D_{\boldsymbol{S_i}})$ based on discretization policy $M = [n_1,n_2,\ldots,n_k]$ in order to compute the likelihood term, $P(D \mid M)$. These components are outlined below.

\subsubsection{Evaluation of $P(D_{\Pi_x} \mid M)$}

Let $J_P = \prod_{i=1}^{n_p} \| P_i \|$. It follows that

\begin{equation}
\label{eq:likelihood_one}
P(D_{\Pi_X} \mid M) = \prod_{i=1}^k  {{1}\over{{n_i + J_P - 1}\choose{J_P -1}}}
{{1}\over{ {{{n}_i}!}\over{ {n^{(p)}_{i,1} !} {n^{(p)}_{i,2} !} \cdots {n^{(p)}_{i,J_P} !}}  }}\text{,}
\end{equation}

\noindent
where $n^{(p)}_{i,j}$ is the number of instances in the $i$th discretized interval with the $j$th value of $\Pi_X$.
Note that $n_i = \sum_{j=1}^{\| \Pi_x \|} n^{(p)}_{i,j}$.
The two factors on the right hand side comes from the second prior: all distributions of values of $\Pi_X$ in a given interval are equiprobable. According to the fourth prior, the distribution in each interval is independent, so the two factors are multiplied together.

\subsubsection{Evaluation of $P(D_{C_i} \mid M, D_{\boldsymbol{S_i}})$}

For each pair $(C_j, \boldsymbol{S_j})$, let $\|C_j \| = J_j$ and $\| \boldsymbol{S_j} \| = L_j = \prod_{v \in \boldsymbol{S_j}} \| v \|$. It follows that

\begin{equation}
\label{eq:likelihood_two}
P(D_{C_j}  \mid M, D_{S_j}) =
\prod_{i=1}^{k} \prod_{l=1}^{L_j} {{1}\over{{n_{i,l} + J_j - 1}\choose{J_j-1}}}
{{1}\over{ {{n_{i,l}}!}\over{ {n^{(j)}_{i,1,l} !} {n^{(j)}_{i,2,l} !} \cdots {n^{(j)}_{i,J_j,l} !}}  }}\text{,}
\end{equation}

\noindent
where $n^{(j)}_{i,m,l}$ is the number of instances in $i$th interval with $m$th value of $C_j$ and $l$th value of $\boldsymbol{S_j}$, $n_{i,l} = \sum_{m=1}^{J_j} n^{(j)}_{i,m,l}$, and $n_i = \sum_{l=1}^{L_j} n_{i,l}$.
The two factors on the right hand side come from the third prior: all distribution of values of $C_j$ in a given interval and with a given value of $S_j$ are equiprobable. According to the forth prior, these distributions are independent from each other, and one thus takes their product. If $S_j = \emptyset$, then Equation \ref{eq:likelihood_two} is equivalent to

\begin{equation}
\label{eq:likelihood_three}
P(D_{C_j}  \mid M) =
\prod_{i=1}^{k}  {{1}\over{{n_{i} + J_j - 1}\choose{J_j-1}}}
{{1}\over{ {{n_{i}}!}\over{ {n^{(j)}_{i,1,\emptyset} !} {n^{(j)}_{i,2,\emptyset} !} \cdots {n^{(j)}_{i,J_j,\emptyset} !}}  }}\text{,}
\end{equation}

\noindent
where $n^{(j)}_{i,m,\emptyset}$ is the number of instances in the $i$th interval with the $m$th value of $C_j$, and $\sum_{m=1}^{J_j} n^{(j)}_{i,m,\emptyset} = n_i$.

The objective function can be formulated given equations \ref{eq:p_M}, \ref{eq:p_D_given_M},\ref{eq:likelihood_one}, and \ref{eq:likelihood_three}.
The log-inverse of $P(M) \cdot P(D|M)$ is minimized for computational convenience:

\begin{equation}
\label{eq:opt_prob}
\begin{aligned}
& \sum_{i=1}^{k-1} - \log(1 - {\exp(- L \cdot {{x_{s_i+1} - x_{s_i}}\over{x_N - x_1}})}) +  \sum_{i=1}^{k} L \cdot {{x_{s_{i}} - x_{s_{i-1} + 1}}\over{x_N - x_1}} +\\
&  \sum_{j=1}^{n_c} \sum_{i=1}^{k}  \sum_{l=1}^{L_j} \left[  \log{{n_{i,l} + J_j - 1}\choose{J_j-1}} + \log \left( { {{n_{i,l}}!}\over{ {n^{(j)}_{i,1,l} !} {n^{(j)}_{i,2,l} !} \cdots {n^{(j)}_{i,J_j,l} !}} } \right) \right] + \\
& \sum_{i=1}^k \left[  \log {{n_{i} + J_P - 1}\choose{J_P-1}} + \log \left( { {{n_i}!}\over{ {n^{(p)}_{i,1} !} {n^{(p)}_{i,2} !} \cdots {n^{(p)}_{i,J_p} !}} } \right) \right]\text{.}
\end{aligned}
\end{equation}

\subsection{Algorithm}

This section describes the procedure used to minimize the objective function.
Note that the objective function is cumulative over intervals, thus, if a partition of $X$ into $k$ intervals with lengths $n_1,n_2,\ldots,n_k$ is an optimal discretization policy, then any subinterval is optimal for the corresponding subproblem.
It follows that dynamic programming can be used to solve the optimization problem exactly.

Every midpoint between two nonequal consecutive values in $D_X$ is a candidate discretization edge.
Let ${D'_X} = \brock{x'_1,x'_2,\ldots,x'_M}$ be the set of unique values of $D_X$ sorted in ascending order, where $M$ is the number unique values in $D_X$.
Recall that $D_X$ is also sorted.
Let $b = \brock{b_0,b_1,b_2,\ldots,b_M}$ be an increasing sequence of integers marking the repeated value intervals in $D_X$ such that $b_0 = 0$ and $x_{b_{i-1} + 1} = x_{b_{i-1} + 2} = \ldots = x_{b_i} = x'_i$.
By this definition, the allowable discretization positions are $d_i = \paren{x_{b_{i}} + x_{{b_i}+1}}/2$ for all $i = 1,2,\ldots,M-1$.

Precomutation can be done to reduce runtime.
Compute $h(u,v)$ for each interval $I_q$ starting from $x_{u}$ to $x_{v}$ for all $(u,v)$ satisfying $u \leq v$:

>>>>

\section{Introduction}
\label{intro}
Bayesian networks \citep{Pearl_1988, PGM_2009} are an increasingly popular method for modeling uncertainty and causality in science and engineering. They provide an efficient factorization of the joint probability distribution over a set of random variables. Bayesian networks first emerged from artificial intelligence research and have been applied to a wide variety of problems, ranging from decision-making systems \citep{DMU_2015} to medical diagnoses \citep{Lustgarten_2011}. In most cases, we assume that all random variables in Bayesian networks are discrete, since many Bayesian network learning algorithms are unable to deal with continuous variables efficiently. However, many applications
require the use of continuous variables, such as position and velocity in dynamic systems.

There are two common approaches to extending Bayesian networks to continuous variables. The first is to model the conditional probability density of each continuous variable using specific families of parametric distributions, and then to redesign Bayesian network learning algorithms based on their parameters. One successful example is belief propagation in Gaussian graphical models \citep{Weiss_2011}. For other particle shapes \citep{Ihler_2009} and non-linear functions~\todo{[cite]}, these redesigned algorithms are computationally expensive and do not perform well.
\tim{Not sure what `particle shapes' are. Do you mean other parametric distributions?}

The second approach is discretization.
Automated discretization methods have been studied in machine learning and statistics for many years \citep{Dougherty_1995, Kerber_1992, Holte_1993, Fayyad_1993}, primarily for classification problems.
They search for the best discretization policy of a continuous attribute by considering its interation with the target class variable.
Interations and dependencies between variables in Bayesian networks are more complicated, where discretization requires considering the impact on the variable's Markov blanket.
Prior work exists for discretizing continuous variables in naive Bayesian networks and tree-augmented networks \citep{Fried_naive}, but only a few discretization methods for general Bayesian networks have been proposed \citep{Friedman_1996, Kozlov_1997, Monti_1998, Steck_2007}.

This paper describes a new Bayesian discretization method for continuous variables in Bayesian networks, extending prior work on single-variable discretization methods from \citet{Boulle_2006} and \citet{Lustgarten_2011}.
The proposed method optimizes the discretization policy relative to the network and takes parents, children, and spouse variables into account.
The optimal single-variable discretization method is derived in Section \ref{sec:single_var}, and a second heuristic method is proposed which reduces the exponential runtime to polynomial.
Section \ref{sec:multi_var} covers Bayesian networks with multiple continuous variables and discretization while simultaneously learning network structure.
The paper concludes with a comparison against existing minimum-description length \citep{Friedman_1996} and MODL \citet{Boulle_2006} methods on real-world datasets in Section \ref{sec:experiments}.

@@

\section{Related Work}
\label{sec:related_work}
This section reviews the minimum description length (MDL) principle discretization \citep{Friedman_1996} and MODL discretization \citep{Boulle_2006}.
The former is the predominant discretization method for continuous variables in Bayesian networks, and is compared to our proposed method in Section \ref{sec:experiments}.
The latter is a discretization method for one continuous variable against a discrete target class, which our proposed method generalizes to Bayesian networks.
The asymptotical equivalence between MDL and MODL was examined in \citep{VL_2000}.

\subsection{MDL Principle Discretization}
The MDL principle was first proposed by \cite{MDL_1978}, and states that the best model for a dataset is one which minimizes the amount of information needed to describe it \citep{Grunwald_2009}.
MDL methods trade off goodness-of-fit against model complexity to reduce generalization error.
In the context of Bayesian networks, \cite{Friedman_1996} applied the MDL principle to determine the optimal number of discretization intervals for continuous variables and optimal the positions of their discretization boundaries.
Their approach selects a discretization policy that minimizes the sum of the description lengths of the discretized Bayesian network and the information necessary for recovering the continuous values from the discretized data.
In the single continuous variable case, the objective function is

@@

\noindent
% $I(\boldsymbol{A},\boldsymbol{B}) =  \sum_{\boldsymbol{a},\boldsymbol{b}} \hat{P}_D (\boldsymbol{a},\boldsymbol{b}) \log {{\hat{P}_D (\boldsymbol{a},\boldsymbol{b})}\over{\hat{P}_D (\boldsymbol{a}),\hat{P}_D (\boldsymbol{b})}}$
where $I(\boldsymbol{A},\boldsymbol{B})$ is the mutual information between the discrete variables $\boldsymbol{A}$ and $\boldsymbol{B}$, $\|\boldsymbol{X}\|$ is the cardinality of discrete variable $\boldsymbol{X}$, $X^*_i$ is the discretized version of continuous variable $X_i$, and $H(p) = -p \log(p) - (1-p) \log(1-p)$.
The optimal discretization policy can be found with dynamical programming.
For Bayesian networks with multiple continuous variables, the algorithm is applied to one variable at a time iteratively over all continuous variables.
While iterating, only one variable is treated as continuous and other continuous variables are discretized based on an initial discretization policy or the discretization result from a previous iteration.

The network structure is not always known in advance. Bayesian network structure learning can be extended to distributions including continuous variables by alternating between traditional discrete structure learning and optimal discretization. Starting with some preliminary discretization policy, one applys a structure learning algorithm to identify the locally optimal graph structure. The discretization policy is refined based on the learned network. The cycle is continued until convergence.

% The discretization technique proposed by \citet{Friedman_1996} is the most well-known of these methods. It is based on the minimum description length principle (MDL): the optimal discretization policy minimizes the description length of the Bayesian network and discretization policy. The MDL algorithm has a $\bigo\paren{N^3 + \paren{n_c {v_\text{max}}^{(n_c^p)_\text{max}} + {v_\text{max}}^{n_p}} \cdot N^2}$ runtime on Bayesian networks with a single continuous variable and multiple discrete variables, where $N$ is the number of data samples from the joint distribution, $n_p$ and $n_c$ are the numbers of parent and children variables for the continuous variable, $v_\text{max}$ is the largest cardinality over all variables in the Markov blanket, and  ${(n_c^p)_\text{max}}$ is the largest number of parent variables of the continuous variable's children. The MDL method iterates over all continuous variables until convergence in the case of multiple continuous variables, converging in a few cycles according to our tests of real-world data. Unfortunately, MDL tends to produce discrete variables with low cardinality~\todo{[cite]}.

@@

The MODL method uses dynamic programming to find the optimal discretization policy, and has a $\bigo\paren{N^3 + \|\text{Class}\| \cdot N^2}$ runtime, where $N$ is the of number of data instances. \cite{Lustgarten_2011} suggest several formulations for the prior term $P(\text{Model})$\tim{Did I state this correctly?}. If the prior is chosen to reflect uniform prior probability over discretizations the runtime can be reduced to $\bigo\paren{\|\text{Class}\| \cdot N^2}$.\tim{Is a sacrifice made? It is sub-optimal, right?}

@@

Samples can be drawn from the joint distribution represented by a Bayesian network by forward sampling.
Variables are sampled in topological order such that parental instantiations are always known~\citep[see][chap.~22]{algo_2009}.
Furthermore, the parent\rightarrow child sampling order can be reversed if the marginal probability of child variable $P(X_i)$ is known.\tim{Do we use this later?}
By Bayes' rule,

@@

We have the following four priors for the discretization model $M$ that enable us to evaluate $P(M)$ and $P(D \mid M)$:\\
\begin{enumerate}
\item For a discretization edge locating at $(x_i + x_{i+1})/2$ has probability
\begin{equation}
1 - {\exp(- L \cdot {{x_{i+1} - x_i}\over{x_N - x_1}})},
\end{equation}
where $L$ is the largest cardinality number of discrete variables in $X$'s Markov blanket.
\item For a given interval of $M$, every distribution of $\Pi_X$ is equiprobable.
\item For each pair of $(C_i,\boldsymbol{S_i})$ and a given interval of $M$, every distribution of $C_i$ with a value of $\boldsymbol{S_i}$ is equiprobable.
\item The distributions of each $C_i$ and $\Pi_X$ in each interval of $M$ with each value of $\boldsymbol{S_i}$ are independent from each other.
\end{enumerate}

Owing to these priors, we are able to evaluate $P(M)$ and $P(D \mid M)$. For the former, we have
\begin{equation}
P(M) = \prod_{i=1}^k \{ [1 - {\exp(- L \cdot {{x_{s_i+1} - x_{s_i}}\over{x_N - x_1}})}] \cdot \exp(L \cdot {{x_{s_{i}} - x_{s_{i-1} + 1}}\over{x_N - x_1}}) \},
\end{equation}
where $s_i = \sum_{j=1}^k n_j$ and $s_k = N$, $s_0 = 0$. Before evaluating $P(D \mid M)$, notice that
\begin{equation}
P(D \mid M) \propto P(D_{\Pi_x} \mid M) \cdot \prod_{i = 1}^{n_c} P(D_{C_i} \mid M, D_{\boldsymbol{S_i}}).
\end{equation}
The graph structure allows us to only consider the interations between $X$ and other variables in its Markov blanket and thus leads to the factorization. For example, in Figure 1., the corresponding $P(D \mid M)$ is factorized as
\begin{equation}
P(D_{ P_1,P_2,P_3 } \mid M) \cdot P( D_{ C_1 } \mid M,D_{\boldsymbol{S_1}}) \cdot P(D_{C_2} \mid M,D_{ S_2  }).
\end{equation}

\begin{figure}[ht]
    \begin{tabular}{cc}
      \input{graph1}
    \end{tabular}
  \caption{Factorization of $P(D \mid M)$}
\end{figure}

The concept behind the factorization is similar to sampling in a Bayesian network. Once we have the distribution of $X$ and sample $X$ for it, then we can further sample $\Pi_x$. Since the samples of $P_1,P_2,\ldots,P_{n_p}$ given a value of $X$ are not independent, we can not further factorize $P(D_{\Pi_x} \mid M)$ into $\prod_i P(D_{P_i} \mid M)$. Similarly, once we have samples of $X$ and $\boldsymbol{S_i}$, we can sample $C_i$. Since the sampling processes of child variables are independent from each other, we have the product $\prod_{i = 1}^{n_c} P(D_{C_i} \mid M, D_{\boldsymbol{S_i}})$.\\

For the following part, we evaluate $P(D_{\Pi_x} \mid M)$  and $P(D_{C_i} \mid M, D_{\boldsymbol{S_i}})$ based on discretization policy $M = [n_1,n_2,\ldots,n_k]$ and dataset $D$.
\subsubsection{Evaluate $P(D_{\Pi_x} \mid M)$}
Assume that $J_P = \prod_{i=1}^{n_p} \| P_i \|$. Then we have:
\begin{equation}
P(D_{\Pi_X} \mid M) = \prod_{i=1}^k  {{1}\over{{n_i + J_P - 1}\choose{J_P -1}}}
{{1}\over{ {{{n}_i}!}\over{ {n^{(p)}_{i,1} !} {n^{(p)}_{i,2} !} \cdots {n^{(p)}_{i,J_P} !}}  }},
\end{equation}
where $n^{(p)}_{i,j}$ is the number of instances in $i$th discretized interval with $j$th value of $\Pi_X$. Note that $n_i = \sum_{j=1}^{\| \Pi_x \|} n^{(p)}_{i,j}$. The two factors on RHS comes from the second prior: all distributions of values of $\Pi_X$ in a given interval are equiprobable. According to the forth prior, the distribution in each interval is independent, so we multiply all the two factors together.
\subsubsection{Evaluate $P(D_{C_i} \mid M, D_{\boldsymbol{S_i}})$}
Assume for each pair of $(C_j, \boldsymbol{S_j})$, we have $\|C_j \| = J_j$ and $\| \boldsymbol{S_j} \| = L_j = \prod_{v \in \boldsymbol{S_j}} \| v \|$. Therefore,
\begin{equation}
P(D_{C_j}  \mid M, D_{S_j}) =
\prod_{i=1}^{k} \prod_{l=1}^{L_j} {{1}\over{{n_{i,l} + J_j - 1}\choose{J_j-1}}}
{{1}\over{ {{n_{i,l}}!}\over{ {n^{(j)}_{i,1,l} !} {n^{(j)}_{i,2,l} !} \cdots {n^{(j)}_{i,J_j,l} !}}  }},
\end{equation}
where $n^{(j)}_{i,m,l}$ is the number of instances in $i$th interval with $m$th value of $C_j$ and $l$th value of $\boldsymbol{S_j}$, $n_{i,l} = \sum_{m=1}^{J_j} n^{(j)}_{i,m,l}$, and $n_i = \sum_{l=1}^{L_j} n_{i,l}$. The two factors on RHS comes from the third prior: all distribution of values of $C_j$ in a given interval and with a given value of $S_j$ are equiprobable. According to the forth prior, these distributions are independent from each other, therefore we multiply all factors. If $S_j = \emptyset$, then Equation 13. is equivalent to
\begin{equation}
P(D_{C_j}  \mid M) =
\prod_{i=1}^{k}  {{1}\over{{n_{i} + J_j - 1}\choose{J_j-1}}}
{{1}\over{ {{n_{i}}!}\over{ {n^{(j)}_{i,1,\emptyset} !} {n^{(j)}_{i,2,\emptyset} !} \cdots {n^{(j)}_{i,J_j,\emptyset} !}}  }},
\end{equation}
where $n^{(j)}_{i,m,\emptyset}$ is the number of instances in $i$th interval with $m$th value of $C_j$, and $\sum_{m=1}^{J_j} n^{(j)}_{i,m,\emptyset} = n_i$. \\

With Equation 8,9,11,12, and 13, now we are able to write down the objective function. Instead of maximizing $P(M) \cdot P(D|M)$, we minimize its log-inverse for convenience. The objective function is
\begin{equation}
\begin{aligned}
& \sum_{i=1}^{k-1} - \log(1 - {\exp(- L \cdot {{x_{s_i+1} - x_{s_i}}\over{x_N - x_1}})}) +  \sum_{i=1}^{k} L \cdot {{x_{s_{i}} - x_{s_{i-1} + 1}}\over{x_N - x_1}} +\\
&  \sum_{j=1}^{n_c} \sum_{i=1}^{k}  \sum_{l=1}^{L_j} \left[  \log{{n_{i,l} + J_j - 1}\choose{J_j-1}} + \log \left( { {{n_{i,l}}!}\over{ {n^{(j)}_{i,1,l} !} {n^{(j)}_{i,2,l} !} \cdots {n^{(j)}_{i,J_j,l} !}} } \right) \right] + \\
& \sum_{i=1}^k \left[  \log {{n_{i} + J_P - 1}\choose{J_P-1}} + \log \left( { {{n_i}!}\over{ {n^{(p)}_{i,1} !} {n^{(p)}_{i,2} !} \cdots {n^{(p)}_{i,J_p} !}} } \right) \right].
\end{aligned}
\end{equation}
All parameters and variables in the objective function are explained in previous subsections.
\subsection{Algorithm}
\label{algo}
Once the objective function is established, the next problem is how to find a discretization model that minimizes the objective function. Note that the objective function is cumulative on intervals. Therefore, if a partition of continuous variable $X$ into $k$ intervals with lengths $n_1,n_2,\ldots,n_k$ is an optimal discretization policy, then $\{ n_2, n_3,\ldots,n_k \}$ is optimal for the subproblem, i.e., the last $N - n_1$ instances of the dataset which has been sorted in ascending order of $D_X$. With this property, we can adapt dynamical programming to solve this optimization problem.

If there is no repeated value in $D_X$, all middle points of two consecutive values can a discretization edge. If there exists repeated values in $D_X$, only the middle points of two consecutive and different values can be a discretization edge. Assume there are $M$ unique values in $D_X$, then ${D'_X} = \{ x'_1,x'_2,\ldots,x'_M \}$ is the set of unique values of $D_X$ in ascending order. Since $D_X$ is also sorted, let $b = \{ b_0,b_1,b_2,\ldots,b_M \}$ be an increasing sequence of integers such that $b_0 = 0$ and $x_{b_{i-1} + 1} = x_{b_{i-1} + 2} = \ldots = x_{{b_i}} = x'_i$. By this definition, the allowable discretization positions are $d_i = ( x_{b_{i}} + x_{{b_i}+1})/2$ for all $i = 1,2,\ldots,M-1$.

Before doing the dynamical programming, in order to save runtime, we first calculate the following function $h(u,v)$ for a interval $I_q$ starting from $x_{u}$ to $x_{v}$ with all $u$,$v$ satisfying $u \leq v$:
\begin{equation}
\begin{aligned}
h(u,v) &=  \log {{n_{q} + J_P - 1}\choose{J_P-1}} + \log \left( { {{n_q}!}\over{ {n^{(p)}_{q,1} !} {n^{(p)}_{q,2} !} \cdots {n^{(p)}_{q,J_p} !}} } \right) \\
& + \sum_{j=1}^{n_c} \sum_{l=1}^{L_j} \left[  \log{{n_{q,l} + J_j - 1}\choose{J_j-1}} + \log \left( { {{n_{q,l}}!}\over{ {n^{(j)}_{q,1,l} !} {n^{(j)}_{q,2,l} !} \cdots {n^{(j)}_{q,J_j,l} !}} } \right) \right]
\end{aligned}
\end{equation}
